Automatic Parallelization and Transparent Fault Tolerance

### 2016-07-12 pubs/tfp16-draft/paper.tex
fitting to page limit

\subsection{Strictness and Parallelism}
%It is a trivial observation that given strict functional semantics, wherein
%function arguments may be safely evaluated before function evaluation, it is
%safe to evaluate the arguments, and the function, in parallel, and that this
%principle applies recursively in the (dynamic) expression tree (or graph).

Given strict functional semantics wherein function arguments may be safely
evaluated before function evaluation or call, it is safe to evaluate the
arguments, and the function, in parallel, and that this applies recursively in
the (dynamic) expression tree (or graph), and similarly for data constructors
as functions.

\section{Project and Goals}
Our project is the light-weight implementation of a pure, higher-order,
polymorphic, functional language and runtime system with which we can
experiment with automatic parallelization strategies with varying degrees of
language strictness, and secondarily, with mechanisms for transparent fault
tolerance.  Light-weight has several implications.  First and foremost it
means that \emph{we are not attempting to compete with GHC in any
  respect}---succumbing to the temptation of recreating a significant subset
of GHC language, type system, or runtime features, or significant code
optimization, would simply divert us from our central goals.\footnote{We
  estimate that we will have, to within an order of magnitude, one person-hour
  of total available development time for each person-year that GHC currently
  represents, or in other units, one second for each hour.}
For another, it means that we are not undertaking significant morphing of GHC
itself: because this is in large part an undergraduate student project it must
be kept as simple as reasonably possible.

Our initial requirements are that
\begin{itemize}
\item The implementation must be feasible in terms of available effort;
\item The language must be strongly typed, implicitly or explicitly.  More
  specifically, the language should explicitly support both Hindley-Milner
  polymorphism and unboxed types~\cite{Jones:1991};
\item Straightforward direct linkage with C, preferably without a sophisticated
  foreign function interface, is essential;
  \item The language should be \emph{currying-friendly} as described by Marlow
and Peyton Jones~\cite{Marlow:2004};
\item Choice of degree of strictness should be easily selected;
\item Sharing, in the sense of what \emph{lazy} means vs.\ merely non-strict,
  should be preserved for all degrees of strictness;
\item The generated code is C;
\item A certain consistency in tail calling in C must be achieved.
\end{itemize}
%
For feasibility we restrict ourselves to implicit parallelism on shared-memory
multiprocessors.  The requirement for tail calling is not primarily for
efficiency, rather for an efficient parallel implementation.  Achieving
this, and thereby sufficiently limiting the use of the native C stack, will be
discussed at some length.

@@

\emph{Nihil sub sole novum}---there is nothing new under the sun. % (Ecclesistes 1:9).
%every new idea has some sort of precedent or echo from the past~\cite{nihil}.

@@

\subsubsection{Strict but still pure.}
The concept of strict but still pure in some sense dates back to the concept
of applicative order reduction in the lambda calculus.  In terms of a
notionally concrete programming language the first thorough explications may
be Backus' FP~\cite{Backus:1978} and follow-on FL
implementation~\cite{AikenFL,FLreport89}.
%
A notable argument for the potential merits of a strict-by-default
pure functional language over a non-strict one is by Sheard, who also provided
a very simple embedded implementation~\cite{Sheard:2003}.
%
A plug-in for GHC (prior to 8.0) exists that makes function bodies strict in
their non-recursive \texttt{let} bindings using a transformation on GHC Core
that recursively expands those \texttt{let} bindings into \texttt{case}
bindings~\cite{Bolingbroke:2008,strict-ghc-plugin}.
%
The Disciplined Disciple Compiler project appears to intend implementation of
a strict, higher order, functional language (or family of languages) compiler
and has been underway for some time, but the emphasis appears to be on a much
more complex type system for handling stateful computation, and the prospects
for its eventual completion are unknown~\cite{disciplined-disciple}.

\subsubsection{Strict, pure, auto-parallelizing.}
NVIDIA's NOVA language, and corresponding compiler, implements a strict
(``call by value''), pure, polymorphic, higher-order functional language with
built-in parallel operators (e.g., \emph{map}, \emph{reduce}, and
\emph{scan})~\cite{NVIDIA:2013}. Three back-ends were implemented: for
sequential C, parallel (multi-threaded) C, and CUDA C, the lattermost
targeting NVIDIA GPUs.

\subsubsection{GHC as a front-end.}
The Intel Labs Haskell Research Compiler (HRC) is the most closely related
work to ours of which we are aware~\cite{Liu:2013,Petersen:2013}.  As in our
planned approach they use GHC as the front end, but exit GHC at the final GHC
Core stage, just before transformation to STG.  The essence of their effort is
to target an existing, highly optimizing, Intel functional language compiler
that is ``largely language agnostic'' and performs (among many other
optimizations) SIMD vectorization.

@@

\subsubsection{Fault tolerance for functional programming.}
Pointon et al.\ developed a limited fault tolerance mechanism for Glasgow
distributed Haskell based on distributed exception
handling~\cite{Pointon:2001}.  However, it appears that their scheme for
handling failed processing elements correctly may have
never been implemented~\cite{Trinder:2000}.

@@

\subsubsection{Removal of explicit heap objects.}
In STG, all object definitions require explicit classification with
\texttt{FUN}, \texttt{PAP}, \texttt{CON}, \texttt{THUNK}. This classification can be
inferred from context by the compiler so these keywords are not necessary.
Functions can then be defined in Haskell style with the formal parameters following
the function's identifier on the left side of the equality. With these simple
changes the syntax becomes a subset of Haskell.

@@

\subsubsection{Avoiding building thunks.}
As \emph{EA} points out, there is no need to build a thunk for the scrutinee
of a \texttt{case} expression because if the \texttt{case} is evaluated the
scrutinee is certain to be evaluated.  Similarly, if a function is strict
there is no need to build a thunk for each of its arguments (assuming boxed
argument type).  Because STG allows only \emph{atoms} (variables or literal
values) in argument positions, naive translation of Haskell \texttt{f e} to
\texttt{let \{x = e\} in f x}, followed by naive code generation, will result in
an unnecessary heap allocation.  An alternative is to translate to
\texttt{case e of x \{\_ -> f x\}}, which we regard as a role of a front-end
such as mini-Haskell.

\subsubsection{The layout rule.}
Mini-Haskell implements Haskell's layout rule in an almost direct translation
of the specification given in section 10.3 of the Haskell 2010
Report~\cite{haskell2010report}. The notable exception is the omission of any
attempt to define the \texttt{parse-error} function.  This is apparently a
tricky spot for Haskell compilers (even GHC has not always had it right) so,
in the name of tradition, it remains a tricky spot for our mini-Haskell front
end.  The edge cases that it will incorrectly handle as a result of this are
exactly that: edge cases.  As such, they are easy to avoid in practice.

@@

\subsubsection{Expression to expression application.}

In STG function application must be to atomic values, i.e., either literals or
variables.  This often requires introducing bindings with either \texttt{let} or
\texttt{case} expressions.  Consider the \texttt{append} function that
concatenates its two list parameters, which can be expressed in STG by
{\footnotesize
\begin{verbatim}
   append = FUN(l1 l2 ->
               case l1 of
                 { Nil -> l2;
                   Cons hd tl -> 
                     let { rec = THUNK(append tl l2);
                           result = CON(Cons hd rec) }
                     in result })
\end{verbatim}}
\noindent In mini-Haskell expressions may be applied to other expressions, so the
recursive call to \texttt{append} can be parenthesized. There is no need for the
\texttt{let} expression:
{\footnotesize
\begin{verbatim}
   append l1 l2 = case l1 of
                    Nil -> l2
                    Cons hd tl -> Cons hd (append tl l2)
\end{verbatim}}
\noindent This also serves as a good example of how expression to expression application is
transformed into STG.  Any non-atomic expression in an application expression is
bound to a heap object in a generated \texttt{let} expression.  These
expressions are thus properly ``atomized'' for STG.  The above mini-Haskell code
is transformed into STG that is structurally identical to the STG code that
precedes it.

\subsubsection{Fancy pattern matching.}

STG supports pattern matching but only in a simple form and only in
\texttt{case} expressions.  Mini-Haskell extends this to something approaching
what is found in Haskell.  Functions, including lambda expressions, may match on
all their arguments. Named functions may have multiple definitions to exhaust
the combinations of patterns in their arguments and patterns in all contexts
(functions and case expressions) may be nested arbitrarily deep.  This feature
greatly simplifies the task of defining more complex behavior.  Without it,
each matched parameter of a function requires an additional level of nested
\texttt{case} expressions. The same is true of each level of nesting within a
single pattern.
%
Even relatively simple functions can be painful to write, such as \texttt{unzip}:
{\footnotesize
\begin{verbatim}
   unzip = FUN(list ->
     case list of { Nil -> let {res = CON( TP2 nil nil)} in res;
                    Cons x xs ->
                      case x of {
                        TP2 a b ->     --TP2 is the pair constructor
                          case unzip xs of {
                            TP2 as bs ->
                              let {aas = CON(Cons a as);
                                   bbs = CON(Cons b bs);
                                   res = CON(TP2 aas bbs)}
                              in res }}})
\end{verbatim}}
\noindent In mini-Haskell this becomes
{\footnotesize
\begin{verbatim}
   unzip :: List (Tupl2 a b) -> Tupl2 (List a) (List b)
   unzip Nil = TP2 Nil Nil
   unzip (Cons (TP2 a b) xs) =
     case unzip xs of
       TP2 as bs -> TP2 (Cons a as) (Cons b bs)
\end{verbatim}}

@@

\subsubsection{Less significant features.}

Mini-Haskell provides some smaller conveniences that don't warrant individual
sections but are worth mentioning.
\begin{itemize}
\item \emph{Constructors as functions.}
  As in Haskell, constructors are first class, acting as true functions.  This
  is implemented by simply generating a function that wraps the appropriate
  \texttt{let}-bound \texttt{CON} object in STG.  Any partially applied
  constructor generates a call to this function.  Fully applied constructors
  generate local \texttt{let}-bound objects to avoid an unnecessary function
  call.

>>>>

%\subsection{Strictness and Parallelism}
%It is a trivial observation that given strict functional semantics, wherein
%function arguments may be safely evaluated before function evaluation, it is
%safe to evaluate the arguments, and the function, in parallel, and that this
%principle applies recursively in the (dynamic) expression tree (or graph).

\section{Project and Goals}
Given strict functional semantics wherein function arguments may be safely
evaluated before function evaluation or call, it is safe to evaluate the
arguments, and the function, in parallel, and that this applies recursively in
the (dynamic) expression tree (or graph), and similarly for data constructors
as functions.
%
Our project is the light-weight implementation of a pure, higher-order,
polymorphic, functional language and runtime system with which we can
experiment with automatic parallelization strategies with varying degrees of
language strictness, and secondarily, with mechanisms for transparent fault
tolerance.  Light-weight has several implications.  First and foremost it
means that \emph{we are not attempting to compete with GHC in any
  respect}---succumbing to the temptation of recreating a significant subset
of GHC language, type system, or runtime features, or significant code
optimization, would simply divert us from our central goals.\footnote{We
  estimate that we will have, to within an order of magnitude, one person-hour
  of total available development time for each person-year that GHC currently
  represents, or in other units, one second for each hour.}
For another, it means that we are not undertaking significant morphing of GHC
itself: because this is in large part an undergraduate student project it must
be kept as simple as reasonably possible.
%
Our initial requirements are that
\begin{itemize}
\item The implementation must be feasible in terms of available effort;
\item The language must be strongly typed, implicitly or explicitly.  More
  specifically, the language should explicitly support both Hindley-Milner
  polymorphism and unboxed types~\cite{Jones:1991};
\item Straightforward direct linkage with C, preferably without a sophisticated
  foreign function interface, is essential;
  \item The language should be \emph{currying-friendly} as described by Marlow
and Peyton Jones~\cite{Marlow:2004};
\item Choice of degree of strictness should be easily selected;
\item Sharing, in the sense of what \emph{lazy} means vs.\ merely non-strict,
  should be preserved for all degrees of strictness;
\item The generated code is C;
\item A certain consistency in tail calling in C must be achieved.
\end{itemize}
%
For feasibility we restrict ourselves to implicit parallelism on shared-memory
multiprocessors.  The requirement for tail calling is not primarily for
efficiency, rather for an efficient parallel implementation.  Achieving
this, and thereby sufficiently limiting the use of the native C stack, will be
discussed at some length.

@@

%\emph{Nihil sub sole novum}---there is nothing new under the sun. % (Ecclesistes 1:9).
%every new idea has some sort of precedent or echo from the past~\cite{nihil}.
There is a vast body of related work; we highlight only the very most closely related.

@@

\subsubsection{Strict but still pure.}
\vspace{-0.1in}
%\paragraph{Strict but still pure.}
The concept of strict but still pure in some sense dates back to the concept
of applicative order reduction in the lambda calculus.  In terms of a
notionally concrete programming language the first thorough explications may
be Backus' FP~\cite{Backus:1978} and follow-on FL
implementation~\cite{AikenFL,FLreport89}.
%
A notable argument for the potential merits of a strict-by-default
pure functional language over a non-strict one is by Sheard, who also provided
a very simple embedded implementation~\cite{Sheard:2003}.
%
A plug-in for GHC (prior to 8.0) exists that makes function bodies strict in
their non-recursive \texttt{let} bindings using a transformation on GHC Core
that recursively expands those \texttt{let} bindings into \texttt{case}
bindings~\cite{Bolingbroke:2008,strict-ghc-plugin}.
%
The Disciplined Disciple Compiler project appears to intend implementation of
a strict, higher order, functional language (or family of languages) compiler
and has been underway for some time, but the emphasis appears to be on a much
more complex type system for handling stateful computation, and the prospects
for its eventual completion are unknown~\cite{disciplined-disciple}.

\vspace{-0.1in}
\subsubsection{Strict, pure, auto-parallelizing.}
%\paragraph{Strict, pure, auto-parallelizing.}
NVIDIA's NOVA language, and corresponding compiler, implements a strict
(``call by value''), pure, polymorphic, higher-order functional language with
built-in parallel operators (e.g., \emph{map}, \emph{reduce}, and
\emph{scan})~\cite{NVIDIA:2013}. Three back-ends were implemented: for
sequential C, parallel (multi-threaded) C, and CUDA C, the lattermost
targeting NVIDIA GPUs.

\vspace{-0.1in}
\subsubsection{GHC as a front-end.}
%\paragraph{GHC as a front-end.}
The Intel Labs Haskell Research Compiler (HRC) is the most closely related
work to ours of which we are aware~\cite{Liu:2013,Petersen:2013}.  As in our
planned approach they use GHC as the front end, but exit GHC at the final GHC
Core stage, just before transformation to STG.  The essence of their effort is
to target an existing, highly optimizing, Intel functional language compiler
that is ``largely language agnostic'' and performs (among many other
optimizations) SIMD vectorization.

@@

\vspace{-0.1in}
\subsubsection{Fault tolerance for functional programming.}
%\paragraph{Fault tolerance for functional programming.}
Pointon et al.\ developed a limited fault tolerance mechanism for Glasgow
distributed Haskell based on distributed exception
handling~\cite{Pointon:2001}.  However, it appears that their scheme for
handling failed processing elements correctly may have
never been implemented~\cite{Trinder:2000}.

@@

\vspace{-0.1in}
\subsubsection{Removal of explicit heap objects.}
In STG, all object definitions require explicit classification with
\texttt{FUN}, \texttt{PAP}, \texttt{CON}, \texttt{THUNK}. This classification can be
inferred from context by the compiler so these keywords are not necessary.
Functions can then be defined in Haskell style with the formal parameters following
the function's identifier on the left side of the equality. With these simple
changes the syntax becomes a subset of Haskell.

@@

\vspace{-0.1in}
\subsubsection{Avoiding building thunks.}
As \emph{EA} points out, there is no need to build a thunk for the scrutinee
of a \texttt{case} expression because if the \texttt{case} is evaluated the
scrutinee is certain to be evaluated.  Similarly, if a function is strict
there is no need to build a thunk for each of its arguments (assuming boxed
argument type).  Because STG allows only \emph{atoms} (variables or literal
values) in argument positions, naive translation of Haskell \texttt{f e} to
\texttt{let \{x = e\} in f x}, followed by naive code generation, will result in
an unnecessary heap allocation.  An alternative is to translate to
\texttt{case e of x \{\_ -> f x\}}, which we regard as a role of a front-end
such as mini-Haskell.

\vspace{-0.1in}
\subsubsection{The layout rule.}
Mini-Haskell implements Haskell's layout rule in an almost direct translation
of the specification given in section 10.3 of the Haskell 2010
Report~\cite{haskell2010report}. The notable exception is the omission of any
attempt to define the \texttt{parse-error} function.  This is apparently a
tricky spot for Haskell compilers (even GHC has not always had it right) so,
in the name of tradition, it remains a tricky spot for our mini-Haskell front
end.  The edge cases that it will incorrectly handle as a result of this are
exactly that: edge cases.  As such, they are easy to avoid in practice.

@@

\vspace{-0.1in}
\subsubsection{Expression to expression application.}
In STG function application must be to atomic values, i.e., either literals or
variables.  This often requires introducing bindings with either \texttt{let} or
\texttt{case} expressions.  Consider the \texttt{append} function that
concatenates its two list parameters, which can be expressed in STG by
{\footnotesize
\begin{verbatim}
   append = FUN(l1 l2 ->
               case l1 of
                 { Nil -> l2;
                   Cons hd tl -> let { rec = THUNK(append tl l2);
                                       result = CON(Cons hd rec) }
                                 in result })
\end{verbatim}}
\noindent In mini-Haskell expressions may be applied to other expressions, so the
recursive call to \texttt{append} can be parenthesized. There is no need for the
\texttt{let} expression:
{\footnotesize
\begin{verbatim}
   append l1 l2 = case l1 of
                    Nil -> l2
                    Cons hd tl -> Cons hd (append tl l2)
\end{verbatim}}
\noindent This also serves as a good example of how expression to expression application is
transformed into STG.  Any non-atomic expression in an application expression is
bound to a heap object in a generated \texttt{let} expression.  These
expressions are thus properly ``atomized'' for STG.  The above mini-Haskell code
is transformed into STG that is structurally identical to the STG code that
precedes it.

\vspace{-0.1in}
\subsubsection{Fancy pattern matching.}
STG supports pattern matching but only in a simple form and only in
\texttt{case} expressions.  Mini-Haskell extends this to something approaching
what is found in Haskell.  Functions, including lambda expressions, may match on
all their arguments. Named functions may have multiple definitions to exhaust
the combinations of patterns in their arguments and patterns in all contexts
(functions and case expressions) may be nested arbitrarily deep.  This feature
greatly simplifies the task of defining more complex behavior.  Without it,
each matched parameter of a function requires an additional level of nested
\texttt{case} expressions. The same is true of each level of nesting within a
single pattern.
%
Even relatively simple functions can be painful to write, such as \texttt{unzip}:
{\footnotesize
\begin{verbatim}
   unzip = FUN(list ->
     case list of { Nil -> let {res = CON( TP2 nil nil)} in res;
                    Cons x xs ->
                      case x of {
                        TP2 a b ->     --TP2 is the pair constructor
                          case unzip xs of {
                            TP2 as bs ->
                              let {aas = CON(Cons a as);
                                   bbs = CON(Cons b bs);
                                   res = CON(TP2 aas bbs)}
                              in res }}})
\end{verbatim}}
\noindent In mini-Haskell this becomes
{\footnotesize
\begin{verbatim}
   unzip :: List (Tupl2 a b) -> Tupl2 (List a) (List b)
   unzip Nil = TP2 Nil Nil
   unzip (Cons (TP2 a b) xs) =
     case unzip xs of
       TP2 as bs -> TP2 (Cons a as) (Cons b bs)
\end{verbatim}}

@@

\vspace{-0.1in}
\subsubsection{Less significant features.}
Mini-Haskell provides some smaller conveniences that don't warrant individual
sections but are worth mentioning.
\begin{itemize}
\item \emph{Constructors as functions.}
  As in Haskell, constructors are first class, acting as true functions.  This
  is implemented by simply generating a function that wraps the appropriate
  \texttt{let}-bound \texttt{CON} object in STG.  Any partially applied
  constructor generates a call to this function.  Fully applied constructors
  generate local \texttt{let}-bound objects to avoid an unnecessary function
  call.
